\documentclass[12pt,a4paper]{article}

% ─── Packages ───
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{enumitem}
\usepackage{float}
\usepackage{natbib}
\usepackage{bm}
\usepackage{mathtools}
\usepackage{nicefrac}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{tcolorbox}
\usepackage{fancybox}
\usepackage{setspace}
\usepackage{caption}

\usetikzlibrary{shapes,arrows.meta,positioning,fit,calc,patterns,decorations.pathreplacing}
\pgfplotsset{compat=1.18}

\tcbuselibrary{theorems,skins,breakable}

\newtheorem{definition}{Definition}[section]
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{remark}{Remark}[section]

% Equation box
\newtcolorbox{eqbox}[1][]{colback=blue!3,colframe=blue!40!black,fonttitle=\bfseries,title=#1,boxrule=0.5pt}

\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\softmax}{softmax}
\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\RMSNorm}{RMSNorm}
\DeclareMathOperator{\LayerNorm}{LayerNorm}
\DeclareMathOperator{\SiLU}{SiLU}
\DeclareMathOperator{\GELU}{GELU}
\DeclareMathOperator{\RoPE}{RoPE}
\DeclareMathOperator{\sign}{sign}

\hypersetup{
    colorlinks=true,
    linkcolor=blue!70!black,
    citecolor=green!50!black,
    urlcolor=blue!60!black
}

\lstset{
    basicstyle=\ttfamily\small,
    breaklines=true,
    frame=single,
    backgroundcolor=\color{gray!5},
    keywordstyle=\color{blue!70!black}\bfseries,
    commentstyle=\color{green!50!black}\itshape,
    numbers=left,
    numberstyle=\tiny\color{gray},
    numbersep=5pt,
    tabsize=4
}

\onehalfspacing

% ─── Title ───
\title{
    \vspace{-1cm}
    {\Huge\textbf{QRAF}}\\[0.3em]
    {\Large\textbf{Quantized Runtime Artifact Format}}\\[0.8em]
    {\large A High-Performance Local LLM Inference Runtime\\with a Custom Binary Model Format\\Optimized for Apple Silicon}\\[0.8em]
    {\normalsize Technical White Paper --- Version 1.0}
}

\author{
    \textbf{Simon-Pierre Boucher}\\[0.3em]
    D\'{e}partement de finance, assurance et immobilier\\
    Universit\'{e} Laval\\
    Qu\'{e}bec, QC, Canada\\[0.3em]
    \texttt{spbou4@ulaval.ca}\\[0.5em]
    \url{https://github.com/simonpierreboucher02/qraf}
}

\date{April 2026}

\begin{document}

\maketitle
\thispagestyle{empty}

\begin{abstract}
\noindent
We present \textbf{QRAF} (Quantized Runtime Artifact Format), a complete local large language model (LLM) inference runtime implemented from scratch in modern C++17. QRAF introduces a novel binary model format designed for memory-mapped, zero-copy tensor access with native quantization support, paired with an inference engine optimized for Apple Silicon processors (M1--M5). The system achieves a \textbf{41$\times$ speedup} over a scalar baseline through Apple Accelerate framework integration, ARM NEON SIMD kernels, Metal GPU compute shaders, and Grand Central Dispatch (GCD) multi-threading. QRAF supports 7 model architectures (LLaMA, Qwen2, GPT-2, GPT-NeoX, OPT, CodeGen, StarCoder), includes a universal model converter supporting HuggingFace, GGUF, safetensors, and PyTorch formats, and implements LoRA adapter merging, KV cache quantization, speculative decoding, and continuous batching. We provide formal analysis of computational complexity, quantization error bounds, and throughput models. Evaluation on 26 models ranging from 70M to 3B parameters demonstrates inference speeds from 8.5 to 512 tokens per second on consumer Apple Silicon hardware.
\end{abstract}

\vspace{0.5em}
\noindent\textbf{Keywords:} LLM inference, model format, quantization, Apple Silicon, NEON SIMD, Metal GPU, transformer, local AI, mmap, zero-copy

\newpage
\tableofcontents
\newpage

% ═══════════════════════════════════════════════════════════
% SECTION 1: INTRODUCTION
% ═══════════════════════════════════════════════════════════
\section{Introduction}
\label{sec:introduction}

The deployment of large language models (LLMs) has traditionally required cloud infrastructure with specialized accelerators \citep{brown2020language, touvron2023llama}. However, growing concerns about privacy, latency, cost, and offline availability have driven significant interest in \emph{local inference} --- running LLMs directly on consumer hardware \citep{dettmers2022gptq, frantar2023gptq}.

Existing local inference solutions such as llama.cpp \citep{llamacpp2023}, Ollama \citep{ollama2024}, and vLLM \citep{kwon2023vllm} have demonstrated the viability of this approach. However, these systems face several fundamental limitations:

\begin{enumerate}[leftmargin=*]
    \item \textbf{Format rigidity}: Most systems are tied to a single model format (GGUF for llama.cpp, safetensors for vLLM), limiting interoperability across the ecosystem.
    \item \textbf{Architecture coupling}: Adding support for new model architectures (e.g., transitioning from LLaMA to Qwen2 to GPT-NeoX) often requires deep modifications to the inference engine core.
    \item \textbf{Platform specificity}: Few systems fully exploit Apple Silicon's unique heterogeneous hardware stack comprising the Accelerate framework, Metal GPU compute, ARM NEON SIMD, and Grand Central Dispatch threading.
    \item \textbf{Deployment complexity}: Users must navigate format conversion, quantization selection, tokenizer configuration, and system setup manually.
\end{enumerate}

\subsection{Contributions}

We address these limitations with QRAF, making the following contributions:

\begin{enumerate}[leftmargin=*]
    \item A \textbf{novel binary model format} (.qraf) designed for mmap-first, zero-copy tensor access with 64-byte SIMD alignment and native quantization metadata (Section~\ref{sec:format}).
    \item A \textbf{multi-architecture inference engine} supporting 7 transformer families through architecture-aware forward pass dispatch with shared KV cache infrastructure (Section~\ref{sec:engine}).
    \item \textbf{Apple Silicon-specific optimizations} achieving $41\times$ speedup through systematic exploitation of Accelerate, NEON, Metal, and GCD (Section~\ref{sec:optimization}).
    \item \textbf{Formal mathematical analysis} of computational complexity, memory requirements, quantization error bounds, and throughput models (Section~\ref{sec:math}).
    \item \textbf{Advanced inference features}: LoRA adapter merging, KV cache quantization, speculative decoding, and continuous batching with formal guarantees (Section~\ref{sec:advanced}).
    \item Comprehensive evaluation on \textbf{26 models} from 70M to 3B parameters with text quality assessment (Section~\ref{sec:evaluation}).
\end{enumerate}

\subsection{Paper Organization}

The remainder of this paper is organized as follows. Section~\ref{sec:related} reviews related work. Section~\ref{sec:math} establishes the mathematical foundations. Section~\ref{sec:format} describes the QRAF binary format. Section~\ref{sec:engine} presents the multi-architecture inference engine. Section~\ref{sec:optimization} details Apple Silicon optimizations. Section~\ref{sec:advanced} covers advanced features. Section~\ref{sec:converter} describes the universal converter. Section~\ref{sec:evaluation} presents experimental evaluation. Section~\ref{sec:discussion} discusses limitations and trade-offs. Section~\ref{sec:conclusion} concludes.

% ═══════════════════════════════════════════════════════════
% SECTION 2: RELATED WORK
% ═══════════════════════════════════════════════════════════
\section{Related Work}
\label{sec:related}

\subsection{Local LLM Inference Engines}

The landscape of local LLM inference has been shaped by several systems. \textbf{llama.cpp} \citep{llamacpp2023} pioneered efficient CPU-based inference using integer quantization and the GGUF format, achieving remarkable performance through hand-optimized kernels. \textbf{Ollama} \citep{ollama2024} built upon llama.cpp to provide a user-friendly Docker-like interface. \textbf{vLLM} \citep{kwon2023vllm} introduced PagedAttention for efficient GPU memory management. \textbf{MLC-LLM} \citep{mlcllm2023} uses Apache TVM compilation for cross-platform deployment.

\subsection{Model Formats}

\textbf{GGUF} \citep{gguf2023} stores quantized tensors with key-value metadata but couples tensor layout to specific quantization schemes. \textbf{safetensors} \citep{safetensors2023} provides safe tensor storage without Python pickle but lacks quantization metadata and tokenizer embedding. \textbf{PyTorch checkpoints} use Python pickle, creating security concerns \citep{pytorch2019}.

\subsection{Quantization Methods}

GPTQ \citep{frantar2023gptq} uses approximate second-order information for optimal rounding. AWQ \citep{lin2024awq} identifies and protects salient weights. SqueezeLLM \citep{kim2024squeezellm} introduces non-uniform quantization with lookup tables.

% ═══════════════════════════════════════════════════════════
% SECTION 3: MATHEMATICAL FOUNDATIONS
% ═══════════════════════════════════════════════════════════
\section{Mathematical Foundations}
\label{sec:math}

This section establishes the mathematical framework underlying the QRAF inference engine.

\subsection{The Transformer Architecture}

\begin{definition}[Decoder-Only Transformer]
\label{def:transformer}
A decoder-only transformer $\mathcal{T}$ is parameterized by:
\begin{itemize}[nosep]
    \item Number of layers $L \in \mathbb{N}$
    \item Hidden dimension $d \in \mathbb{N}$
    \item Number of attention heads $H \in \mathbb{N}$, with head dimension $d_h = d / H$
    \item Number of key-value heads $H_{kv} \in \mathbb{N}$, where $H_{kv} | H$
    \item Feed-forward intermediate dimension $d_{ff} \in \mathbb{N}$
    \item Vocabulary size $V \in \mathbb{N}$
    \item Maximum sequence length $T_{\max} \in \mathbb{N}$
\end{itemize}
The model defines the mapping $f_\theta: \mathbb{N}^{T} \to \mathbb{R}^{V}$ that maps a sequence of $T$ token indices to a logit vector over the vocabulary.
\end{definition}

\subsection{Attention Mechanism}

The core of the transformer is the scaled dot-product attention mechanism \citep{vaswani2017attention}.

\begin{eqbox}[Scaled Dot-Product Attention]
For query $\bm{q} \in \mathbb{R}^{d_h}$, key matrix $\bm{K} \in \mathbb{R}^{T \times d_h}$, and value matrix $\bm{V} \in \mathbb{R}^{T \times d_h}$:
\begin{equation}
\text{Attention}(\bm{q}, \bm{K}, \bm{V}) = \softmax\left(\frac{\bm{q}\bm{K}^\top}{\sqrt{d_h}}\right) \bm{V}
\label{eq:attention}
\end{equation}
\end{eqbox}

The softmax function with causal masking is defined as:
\begin{equation}
\softmax(\bm{z})_i = \frac{\exp(z_i)}{\sum_{j=1}^{T} \exp(z_j)}, \quad \text{for } i \leq t \text{ (causal)}
\label{eq:softmax}
\end{equation}

For numerical stability, we compute:
\begin{equation}
\softmax(\bm{z})_i = \frac{\exp(z_i - \max_j z_j)}{\sum_{j} \exp(z_j - \max_j z_j)}
\label{eq:softmax_stable}
\end{equation}

\subsubsection{Multi-Head Attention}

Given input $\bm{x} \in \mathbb{R}^d$ at position $t$, the multi-head attention computes:

\begin{align}
\bm{q}_h &= \bm{W}_Q^{(h)} \bm{x} + \bm{b}_Q^{(h)} \in \mathbb{R}^{d_h}, \quad h = 1, \ldots, H \label{eq:q_proj} \\
\bm{k}_g &= \bm{W}_K^{(g)} \bm{x} + \bm{b}_K^{(g)} \in \mathbb{R}^{d_h}, \quad g = 1, \ldots, H_{kv} \label{eq:k_proj} \\
\bm{v}_g &= \bm{W}_V^{(g)} \bm{x} + \bm{b}_V^{(g)} \in \mathbb{R}^{d_h} \label{eq:v_proj}
\end{align}

where $\bm{W}_Q^{(h)} \in \mathbb{R}^{d_h \times d}$, and bias terms are optional (present in Qwen2, GPT-2; absent in LLaMA).

\subsubsection{Grouped Query Attention}

\begin{definition}[Grouped Query Attention (GQA) \citep{ainslie2023gqa}]
\label{def:gqa}
In GQA, each key-value head $g$ is shared by $r = H / H_{kv}$ query heads. For query head $h$, the corresponding KV head index is $g(h) = \lfloor h / r \rfloor$.
\end{definition}

The attention weights for head $h$ at position $t$ attending to position $s \leq t$:

\begin{eqbox}[GQA Attention Score]
\begin{equation}
\alpha_{h,t,s} = \frac{\exp\left(\frac{\bm{q}_{h,t}^\top \bm{k}_{g(h),s}}{\sqrt{d_h}}\right)}{\displaystyle\sum_{s'=1}^{t} \exp\left(\frac{\bm{q}_{h,t}^\top \bm{k}_{g(h),s'}}{\sqrt{d_h}}\right)}
\label{eq:gqa_score}
\end{equation}
\end{eqbox}

The per-head output and final multi-head output:
\begin{align}
\bm{o}_{h,t} &= \sum_{s=1}^{t} \alpha_{h,t,s} \cdot \bm{v}_{g(h),s} \label{eq:head_output} \\
\text{MHA}(\bm{x}, t) &= \bm{W}_O \left[\bm{o}_{1,t}; \bm{o}_{2,t}; \ldots; \bm{o}_{H,t}\right] \label{eq:mha_output}
\end{align}

\begin{remark}[Special Cases of GQA]
\begin{itemize}[nosep]
    \item $H_{kv} = H$: Standard Multi-Head Attention (MHA) \citep{vaswani2017attention}
    \item $H_{kv} = 1$: Multi-Query Attention (MQA) \citep{shazeer2019mqa}
    \item $1 < H_{kv} < H$: Grouped Query Attention (GQA) \citep{ainslie2023gqa}
\end{itemize}
\end{remark}

The KV cache memory scales as $O(L \cdot H_{kv} \cdot d_h \cdot T)$ rather than $O(L \cdot H \cdot d_h \cdot T)$. For Qwen2.5-3B with $H=16, H_{kv}=2$, this is an $8\times$ reduction.

% ─── Figure: GQA Diagram ───
\begin{figure}[H]
\centering
\begin{tikzpicture}[
    head/.style={draw, rounded corners, minimum width=1.2cm, minimum height=0.6cm, font=\small},
    kv/.style={draw, rounded corners, fill=blue!15, minimum width=1.2cm, minimum height=0.6cm, font=\small},
    q/.style={draw, rounded corners, fill=orange!20, minimum width=0.8cm, minimum height=0.6cm, font=\small},
    arr/.style={-{Stealth}, thick}
]
    % KV heads
    \node[kv] (kv1) at (0, 0) {KV$_1$};
    \node[kv] (kv2) at (3, 0) {KV$_2$};

    % Q heads for KV1
    \node[q] (q1) at (-1.2, 1.8) {Q$_1$};
    \node[q] (q2) at (-0.4, 1.8) {Q$_2$};
    \node[q] (q3) at (0.4, 1.8) {Q$_3$};
    \node[q] (q4) at (1.2, 1.8) {Q$_4$};

    % Q heads for KV2
    \node[q] (q5) at (1.8, 1.8) {Q$_5$};
    \node[q] (q6) at (2.6, 1.8) {Q$_6$};
    \node[q] (q7) at (3.4, 1.8) {Q$_7$};
    \node[q] (q8) at (4.2, 1.8) {Q$_8$};

    % Arrows
    \foreach \q in {q1,q2,q3,q4}
        \draw[arr, blue!60] (\q) -- (kv1);
    \foreach \q in {q5,q6,q7,q8}
        \draw[arr, blue!60] (\q) -- (kv2);

    % Labels
    \node[font=\footnotesize, above=0.1cm of q2] {Group 1 ($r=4$)};
    \node[font=\footnotesize, above=0.1cm of q7] {Group 2 ($r=4$)};
    \node[font=\small\bfseries, below=0.3cm of kv1, xshift=1.5cm] {GQA: $H=8$, $H_{kv}=2$, $r=4$};
\end{tikzpicture}
\caption{Grouped Query Attention (GQA). Multiple query heads share a single key-value head, reducing KV cache memory by factor $r = H/H_{kv}$.}
\label{fig:gqa}
\end{figure}

\subsection{Rotary Position Embedding (RoPE)}

\begin{definition}[RoPE \citep{su2024roformer}]
\label{def:rope}
For position $t \in \mathbb{N}$ and dimension pair index $i \in \{0, 1, \ldots, d_h/2 - 1\}$, define the frequency:
\begin{equation}
\theta_i = \theta_{\text{base}}^{-2i/d_h}
\label{eq:rope_freq}
\end{equation}
where $\theta_{\text{base}}$ is a hyperparameter (10,000 for LLaMA, 1,000,000 for Qwen2.5). The $2\times2$ rotation matrix is:
\begin{equation}
\bm{R}_t^{(i)} = \begin{pmatrix} \cos(t\theta_i) & -\sin(t\theta_i) \\ \sin(t\theta_i) & \phantom{-}\cos(t\theta_i) \end{pmatrix}
\label{eq:rope_matrix}
\end{equation}
\end{definition}

The full RoPE transformation applied to a $d_h$-dimensional vector:

\begin{eqbox}[Rotary Position Embedding]
\begin{equation}
\RoPE(\bm{q}, t) = \underbrace{\begin{pmatrix} \bm{R}_t^{(0)} & & \\ & \ddots & \\ & & \bm{R}_t^{(d_h/2-1)} \end{pmatrix}}_{\displaystyle \bm{R}_t \in \mathbb{R}^{d_h \times d_h}} \bm{q}
\label{eq:rope_full}
\end{equation}
\end{eqbox}

\begin{theorem}[RoPE Relative Position Property]
\label{thm:rope}
For queries $\tilde{\bm{q}}_m = \bm{R}_m \bm{q}$ and keys $\tilde{\bm{k}}_n = \bm{R}_n \bm{k}$:
\begin{equation}
\tilde{\bm{q}}_m^\top \tilde{\bm{k}}_n = \bm{q}^\top \bm{R}_{n-m} \bm{k}
\label{eq:rope_relative}
\end{equation}
Thus the dot product depends only on the relative position $(n - m)$, not the absolute positions.
\end{theorem}

\begin{proof}
Since each block $\bm{R}_t^{(i)}$ is an orthogonal rotation matrix:
\begin{align}
\tilde{\bm{q}}_m^\top \tilde{\bm{k}}_n &= (\bm{R}_m \bm{q})^\top (\bm{R}_n \bm{k}) = \bm{q}^\top \bm{R}_m^\top \bm{R}_n \bm{k}
\end{align}
For orthogonal matrices, $\bm{R}_m^\top = \bm{R}_m^{-1} = \bm{R}_{-m}$. Therefore:
\begin{equation}
\bm{R}_m^\top \bm{R}_n = \bm{R}_{-m} \bm{R}_n = \bm{R}_{n-m}
\end{equation}
where the last equality follows from the rotation composition property: $\bm{R}_a \bm{R}_b = \bm{R}_{a+b}$, since multiplying rotation matrices corresponds to adding angles. This holds independently for each $2\times2$ block.
\end{proof}

% ─── Figure: RoPE Visualization ───
\begin{figure}[H]
\centering
\begin{tikzpicture}[scale=1.3]
    % Axes
    \draw[-{Stealth}] (-2.5,0) -- (2.5,0) node[right] {$q_{2i}$};
    \draw[-{Stealth}] (0,-2.5) -- (0,2.5) node[above] {$q_{2i+1}$};

    % Original vector
    \draw[-{Stealth}, thick, blue] (0,0) -- (1.8, 0.8) node[right, font=\small] {$\bm{q}$ (pos 0)};

    % Rotated vectors
    \draw[-{Stealth}, thick, red!70!black] (0,0) -- (1.2, 1.6) node[above right, font=\small] {$\bm{R}_1\bm{q}$ (pos 1)};
    \draw[-{Stealth}, thick, green!60!black] (0,0) -- (0.3, 1.95) node[above, font=\small] {$\bm{R}_2\bm{q}$ (pos 2)};
    \draw[-{Stealth}, thick, purple] (0,0) -- (-0.7, 1.85) node[above left, font=\small] {$\bm{R}_3\bm{q}$ (pos 3)};

    % Arc
    \draw[dashed, gray] (0,0) circle (1.97);

    % Angle arc
    \draw[-{Stealth}, orange, thick] (1.5,0.67) arc[start angle=24, end angle=53, radius=1.65];
    \node[orange, font=\small] at (1.9, 1.3) {$\theta_i$};
\end{tikzpicture}
\caption{RoPE rotates query/key vectors by an angle proportional to position $t$. The dot product between two rotated vectors depends only on the difference in rotation angles, encoding relative position.}
\label{fig:rope}
\end{figure}

\subsection{Normalization Layers}

\begin{definition}[RMSNorm \citep{zhang2019rmsnorm}]
For input $\bm{x} \in \mathbb{R}^d$ with learnable weight $\bm{\gamma} \in \mathbb{R}^d$:
\begin{equation}
\RMSNorm(\bm{x}) = \frac{\bm{x}}{\sqrt{\frac{1}{d}\displaystyle\sum_{i=1}^{d} x_i^2 + \epsilon}} \odot \bm{\gamma}
\label{eq:rmsnorm}
\end{equation}
\end{definition}

\begin{definition}[LayerNorm \citep{ba2016layernorm}]
\begin{equation}
\LayerNorm(\bm{x}) = \frac{\bm{x} - \mu(\bm{x})}{\sigma(\bm{x})} \odot \bm{\gamma} + \bm{\beta}
\label{eq:layernorm}
\end{equation}
where $\mu(\bm{x}) = \frac{1}{d}\sum_{i} x_i$ and $\sigma(\bm{x}) = \sqrt{\frac{1}{d}\sum_{i}(x_i - \mu)^2 + \epsilon}$.
\end{definition}

\begin{proposition}[Computational Cost Comparison]
\label{prop:norm_cost}
RMSNorm requires $3d + O(1)$ FLOPs: $d$ multiplications for $x_i^2$, $d-1$ additions, 1 division, $d$ multiplications for scaling. LayerNorm requires $5d + O(1)$ FLOPs: additionally $d$ subtractions for centering and $d$ additions for bias.
\end{proposition}

\subsection{Feed-Forward Networks}

\begin{definition}[SwiGLU MLP \citep{shazeer2020glu}]
\begin{equation}
\text{FFN}_{\text{SwiGLU}}(\bm{x}) = \bm{W}_{\downarrow}\left(\underbrace{\SiLU(\bm{W}_{\text{gate}}\bm{x})}_{\text{gate}} \odot \underbrace{\bm{W}_{\uparrow}\bm{x}}_{\text{up}}\right)
\label{eq:swiglu}
\end{equation}
where $\SiLU(x) = x \cdot \sigma(x) = \frac{x}{1 + e^{-x}}$ is the Sigmoid Linear Unit.
\end{definition}

\begin{definition}[Standard MLP]
\begin{equation}
\text{FFN}_{\text{std}}(\bm{x}) = \bm{W}_2 \cdot \phi\left(\bm{W}_1 \bm{x} + \bm{b}_1\right) + \bm{b}_2
\label{eq:standard_mlp}
\end{equation}
where $\phi$ is GELU \citep{hendrycks2016gelu} for GPT-2/Pythia or ReLU for OPT.
\end{definition}

The GELU activation and its fast approximation:
\begin{align}
\GELU(x) &= x \cdot \Phi(x) = \frac{x}{2}\left(1 + \text{erf}\left(\frac{x}{\sqrt{2}}\right)\right) \label{eq:gelu_exact} \\
&\approx \frac{x}{2}\left(1 + \tanh\left(\sqrt{\frac{2}{\pi}}\left(x + 0.044715 x^3\right)\right)\right) \label{eq:gelu_approx}
\end{align}

% ─── Figure: Activation Functions ───
\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=12cm, height=6cm,
    xlabel={$x$}, ylabel={$\phi(x)$},
    xmin=-4, xmax=4, ymin=-1.5, ymax=4,
    legend pos=north west,
    grid=major,
    grid style={gray!20},
    thick
]
    % SiLU
    \addplot[blue, thick, domain=-4:4, samples=200] {x/(1+exp(-x))};
    \addlegendentry{SiLU: $x\sigma(x)$}
    % GELU
    \addplot[red, thick, domain=-4:4, samples=200] {0.5*x*(1+tanh(0.7978845608*(x+0.044715*x^3)))};
    \addlegendentry{GELU (approx)}
    % ReLU
    \addplot[green!60!black, thick, domain=-4:4, samples=200] {max(0,x)};
    \addlegendentry{ReLU: $\max(0,x)$}
\end{axis}
\end{tikzpicture}
\caption{Activation functions used in different architectures. SiLU (LLaMA, Qwen2) and GELU (GPT-2, Pythia) are smooth approximations of ReLU (OPT).}
\label{fig:activations}
\end{figure}

\subsection{Sampling Theory}

The token generation process involves sampling from the output distribution.

\begin{definition}[Temperature-Scaled Distribution]
Given logits $\bm{z} \in \mathbb{R}^V$ and temperature $\tau > 0$:
\begin{equation}
p_\tau(x = i) = \frac{\exp(z_i / \tau)}{\displaystyle\sum_{j=1}^{V} \exp(z_j / \tau)}
\label{eq:temperature}
\end{equation}
\end{definition}

\begin{proposition}[Temperature Limits]
\begin{itemize}[nosep]
    \item As $\tau \to 0^+$: $p_\tau \to \delta_{\argmax_i z_i}$ (greedy/deterministic)
    \item As $\tau \to \infty$: $p_\tau \to \text{Uniform}(V)$ (maximum entropy)
    \item At $\tau = 1$: standard softmax
\end{itemize}
\end{proposition}

\begin{definition}[Top-$k$ Filtering]
Let $\mathcal{S}_k = \{i_1, \ldots, i_k\}$ be the indices of the $k$ largest logits, sorted such that $z_{i_1} \geq z_{i_2} \geq \cdots \geq z_{i_k}$. The filtered distribution:
\begin{equation}
p_{\text{top-}k}(x=i) = \frac{p_\tau(x=i) \cdot \mathbb{1}[i \in \mathcal{S}_k]}{\displaystyle\sum_{j \in \mathcal{S}_k} p_\tau(x=j)}
\label{eq:topk}
\end{equation}
\end{definition}

\begin{definition}[Nucleus Sampling (Top-$p$) \citep{holtzman2020nucleus}]
The nucleus $\mathcal{N}_p$ is the smallest set such that:
\begin{equation}
\sum_{i \in \mathcal{N}_p} p_\tau(x = i) \geq p
\label{eq:nucleus_def}
\end{equation}
The distribution is renormalized over $\mathcal{N}_p$.
\end{definition}

\begin{definition}[Repetition Penalty \citep{keskar2019ctrl}]
For tokens $i$ in the recent context $\mathcal{C}$, with penalty factor $\rho > 1$:
\begin{equation}
z_i' = \begin{cases}
z_i / \rho & \text{if } z_i > 0 \text{ and } i \in \mathcal{C} \\
z_i \cdot \rho & \text{if } z_i \leq 0 \text{ and } i \in \mathcal{C} \\
z_i & \text{if } i \notin \mathcal{C}
\end{cases}
\label{eq:rep_penalty}
\end{equation}
This reduces the probability of recently generated tokens regardless of their logit sign.
\end{definition}

% ═══════════════════════════════════════════════════════════
% SECTION 4: COMPUTATIONAL COMPLEXITY
% ═══════════════════════════════════════════════════════════
\section{Computational Complexity and Memory Analysis}
\label{sec:complexity}

\subsection{Per-Token FLOPs}

\begin{theorem}[Single-Token Forward Pass Complexity]
\label{thm:flops}
The floating-point operations for generating one token at sequence position $t$ are:
\begin{equation}
\boxed{
\mathcal{F}(t) = L \cdot \left(\mathcal{F}_{\text{attn}}(t) + \mathcal{F}_{\text{MLP}} + \mathcal{F}_{\text{norm}}\right) + \mathcal{F}_{\text{head}}
}
\label{eq:total_flops}
\end{equation}
where:
\begin{align}
\mathcal{F}_{\text{attn}}(t) &= \underbrace{2d \cdot H d_h}_{\bm{W}_Q \bm{x}} + \underbrace{2d \cdot 2H_{kv} d_h}_{\bm{W}_K, \bm{W}_V} + \underbrace{2H d_h \cdot t}_{\bm{q}^\top\bm{K}} + \underbrace{2H d_h \cdot t}_{\bm{\alpha}^\top\bm{V}} + \underbrace{2d \cdot H d_h}_{\bm{W}_O} \label{eq:attn_flops} \\
\mathcal{F}_{\text{MLP}} &= \begin{cases}
3 \times 2d \cdot d_{ff} = 6d \cdot d_{ff} & \text{(SwiGLU: gate, up, down)} \\
2 \times 2d \cdot d_{ff} = 4d \cdot d_{ff} & \text{(Standard: fc1, fc2)}
\end{cases} \label{eq:mlp_flops} \\
\mathcal{F}_{\text{norm}} &\approx 5d \text{ per norm} \label{eq:norm_flops} \\
\mathcal{F}_{\text{head}} &= 2V \cdot d \label{eq:head_flops}
\end{align}
\end{theorem}

\begin{proof}
Each matrix-vector product $\bm{y} = \bm{W}\bm{x}$ with $\bm{W} \in \mathbb{R}^{m \times n}$ requires $m$ dot products of dimension $n$, each costing $2n - 1 \approx 2n$ FLOPs (multiply-accumulate). Thus the total is $2mn$ FLOPs. The QKV projections: Q costs $2d \cdot Hd_h$, K costs $2d \cdot H_{kv}d_h$, V costs $2d \cdot H_{kv}d_h$. Attention scores: $H$ heads, each computing a dot product of dimension $d_h$ with $t$ cached keys, giving $2Hd_ht$ FLOPs. Value aggregation: weighted sum of $t$ vectors of dimension $d_h$ for $H$ heads, giving $2Hd_ht$ FLOPs. Output projection: $2d \cdot Hd_h = 2d^2$ FLOPs. SwiGLU has three matrices; standard MLP has two.
\end{proof}

\begin{corollary}[Numerical Example: Qwen2.5-3B]
With $L=36$, $d=2048$, $H=16$, $H_{kv}=2$, $d_h=128$, $d_{ff}=11008$, $V=151936$, $t=100$:
\begin{align}
\mathcal{F}_{\text{attn}}(100) &= 2(2048)(2048) + 2(2048)(256) + 2(2)(2048)(128)(100) + 2(2048)(2048) \nonumber \\
&= 8.4\text{M} + 1.0\text{M} + 105\text{M} + 8.4\text{M} = 122.8\text{M} \\
\mathcal{F}_{\text{MLP}} &= 6 \times 2048 \times 11008 = 135.3\text{M} \\
\mathcal{F}(100) &\approx 36 \times (122.8 + 135.3)\text{M} + 2(151936)(2048) \nonumber \\
&= 9291\text{M} + 622\text{M} \approx \textbf{9.9 GFLOPs}
\end{align}
\end{corollary}

\subsection{Memory Bandwidth Analysis}

\begin{theorem}[Arithmetic Intensity]
\label{thm:intensity}
The arithmetic intensity (FLOPs per byte of memory accessed) for autoregressive token generation:
\begin{equation}
I(t) = \frac{\mathcal{F}(t)}{B_{\text{weights}} + B_{\text{KV}}(t) + B_{\text{act}}}
\label{eq:intensity}
\end{equation}
where $B_{\text{weights}} = P \cdot b_w$ is total weight bytes, $B_{\text{KV}}(t) = 2L \cdot H_{kv} \cdot d_h \cdot t \cdot b_{\text{kv}}$ is KV cache bytes, and $b_w, b_{\text{kv}}$ are bytes per element.
\end{theorem}

\begin{proposition}[Memory-Bandwidth Bound Regime]
\label{prop:bandwidth}
For autoregressive generation at early positions ($t \ll T_{\max}$) with f32 weights:
\begin{equation}
I \approx \frac{\mathcal{F}(t)}{4P} \approx \frac{2 \cdot \text{(FLOP ratio)}}{1} < 2 \text{ FLOP/byte}
\end{equation}
Since Apple Silicon M1-M5 achieves $>100$ GFLOP/s compute but only $\sim$60--100 GB/s memory bandwidth, the system is \textbf{memory-bandwidth bound} when $I < I_{\text{ridge}} = \text{FLOP/s} / \text{BW}$.
\end{proposition}

% ─── Figure: Roofline ───
\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=12cm, height=7cm,
    xlabel={Arithmetic Intensity (FLOP/byte)},
    ylabel={Performance (GFLOP/s)},
    xmin=0.01, xmax=100,
    ymin=0.1, ymax=1000,
    xmode=log, ymode=log,
    grid=major,
    grid style={gray!20},
    legend pos=south east,
    thick
]
    % Memory bandwidth ceiling (68 GB/s for M1)
    \addplot[blue, very thick, domain=0.01:2.2] {68*x};

    % Compute ceiling
    \addplot[red, very thick, domain=2.2:100] {150};

    % Ridge point
    \addplot[only marks, mark=*, mark size=3pt, black] coordinates {(2.2, 150)};
    \node[above, font=\small] at (axis cs:2.2, 150) {Ridge point};

    % QRAF operating points
    \addplot[only marks, mark=triangle*, mark size=4pt, green!60!black]
        coordinates {(0.5, 34) (0.8, 54) (1.0, 68)};
    \node[right, font=\scriptsize, green!60!black] at (axis cs:0.5, 34) {3B f32};
    \node[right, font=\scriptsize, green!60!black] at (axis cs:0.8, 54) {0.5B f32};
    \node[right, font=\scriptsize, green!60!black] at (axis cs:1.0, 68) {135M f32};

    % Quantized operating points
    \addplot[only marks, mark=square*, mark size=3pt, orange]
        coordinates {(2.0, 130) (4.0, 150)};
    \node[right, font=\scriptsize, orange] at (axis cs:2.0, 130) {Q8};
    \node[right, font=\scriptsize, orange] at (axis cs:4.0, 150) {Q4};

    \legend{Memory BW ceiling, Compute ceiling}
\end{axis}
\end{tikzpicture}
\caption{Roofline model for QRAF on Apple M1. f32 inference operates in the memory-bandwidth bound regime (left of ridge point). Quantization shifts operating points rightward toward compute-bound.}
\label{fig:roofline}
\end{figure}

\subsection{Memory Requirements}

\begin{theorem}[Total Memory Footprint]
\label{thm:memory}
\begin{equation}
\boxed{
M_{\text{total}} = \underbrace{P \cdot \frac{b_w}{8}}_{\text{weights}} + \underbrace{2L \cdot H_{kv} \cdot d_h \cdot T \cdot \frac{b_{\text{kv}}}{8}}_{\text{KV cache}} + \underbrace{O(d + d_{ff})}_{\text{activations}}
}
\label{eq:memory}
\end{equation}
\end{theorem}

\begin{table}[H]
\centering
\caption{Memory breakdown for Qwen2.5-3B at different precisions.}
\label{tab:memory}
\begin{tabular}{@{}lrrrr@{}}
\toprule
\textbf{Component} & \textbf{f32} & \textbf{f16} & \textbf{Q8} & \textbf{Q4} \\
\midrule
Weights ($P = 3\times10^9$) & 12.0 GB & 6.0 GB & 3.0 GB & 1.5 GB \\
KV Cache ($T=2048$) & 151 MB & 75 MB & 38 MB & 19 MB \\
Activations & $\sim$1 MB & $\sim$1 MB & $\sim$1 MB & $\sim$1 MB \\
\midrule
\textbf{Total} & \textbf{12.15 GB} & \textbf{6.07 GB} & \textbf{3.04 GB} & \textbf{1.52 GB} \\
\bottomrule
\end{tabular}
\end{table}

% ═══════════════════════════════════════════════════════════
% SECTION 5: QUANTIZATION THEORY
% ═══════════════════════════════════════════════════════════
\section{Quantization Theory}
\label{sec:quantmath}

\subsection{Block-wise Symmetric Quantization}

\begin{definition}[Q$b$\_0 Block Quantization]
\label{def:quant}
For a block of $B$ floating-point values $\bm{x} = (x_1, \ldots, x_B) \in \mathbb{R}^B$, the $b$-bit symmetric quantization with block size $B$ computes:
\begin{align}
s &= \frac{\|\bm{x}\|_\infty}{2^{b-1} - 1} \label{eq:quant_scale} \\
\hat{x}_i &= \text{clamp}\left(\text{round}\left(\frac{x_i}{s}\right), -(2^{b-1}-1), 2^{b-1}-1\right) \label{eq:quant_round}
\end{align}
The dequantized value is $\tilde{x}_i = s \cdot \hat{x}_i$.
\end{definition}

\begin{theorem}[Quantization Error Bound]
\label{thm:quant_error}
For Q$b$\_0 quantization, the per-element error is bounded by:
\begin{equation}
\boxed{|x_i - \tilde{x}_i| \leq \frac{s}{2} = \frac{\|\bm{x}\|_\infty}{2(2^{b-1} - 1)}}
\label{eq:quant_bound}
\end{equation}
\end{theorem}

\begin{proof}
The rounding operation satisfies $|\text{round}(y) - y| \leq \frac{1}{2}$ for any $y \in \mathbb{R}$. Therefore:
\begin{equation}
|x_i - s \cdot \hat{x}_i| = s \cdot \left|\frac{x_i}{s} - \hat{x}_i\right| = s \cdot \left|\frac{x_i}{s} - \text{round}\left(\frac{x_i}{s}\right)\right| \leq \frac{s}{2}
\end{equation}
Substituting $s = \|\bm{x}\|_\infty / (2^{b-1}-1)$ yields the result.
\end{proof}

\begin{corollary}[Relative Error]
The maximum relative error for non-zero elements:
\begin{equation}
\frac{|x_i - \tilde{x}_i|}{|x_i|} \leq \frac{\|\bm{x}\|_\infty}{2(2^{b-1}-1)|x_i|}
\end{equation}
For the largest element: relative error $\leq \frac{1}{2(2^{b-1}-1)}$, which is $\approx 0.39\%$ for Q8 and $\approx 7.1\%$ for Q4.
\end{corollary}

\subsection{Error Propagation in Matrix-Vector Products}

\begin{theorem}[Quantized MatVec Error]
\label{thm:matvec_error}
For $\bm{y} = \bm{W}\bm{x}$ where $\bm{W} \in \mathbb{R}^{m \times n}$ is quantized row-wise with block size $B$, the output error is:
\begin{equation}
\|\bm{y} - \tilde{\bm{y}}\|_2 \leq \frac{\sqrt{n}}{2(2^{b-1}-1)} \cdot \max_{\text{blocks}} \|\bm{W}_{\text{block}}\|_\infty \cdot \|\bm{x}\|_2
\label{eq:matvec_error}
\end{equation}
\end{theorem}

\begin{proof}
For each output element $y_i = \bm{w}_i^\top \bm{x}$:
\begin{align}
|y_i - \tilde{y}_i| &= |(\bm{w}_i - \tilde{\bm{w}}_i)^\top \bm{x}| \leq \|\bm{w}_i - \tilde{\bm{w}}_i\|_2 \cdot \|\bm{x}\|_2
\end{align}
by Cauchy-Schwarz. The quantization error per block of size $B$ has $\|\bm{e}_b\|_2 \leq \frac{s_b}{2}\sqrt{B}$. Summing over $n/B$ blocks: $\|\bm{w}_i - \tilde{\bm{w}}_i\|_2 \leq \frac{\sqrt{n}}{2(2^{b-1}-1)} \max_b \|\bm{w}_{i,b}\|_\infty$. Taking the $\ell_2$ norm over all $m$ outputs completes the bound.
\end{proof}

% ─── Figure: Quantization Error Distribution ───
\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=12cm, height=6cm,
    xlabel={Weight Value},
    ylabel={Quantized Level},
    xmin=-1.2, xmax=1.2, ymin=-1.2, ymax=1.2,
    grid=major, grid style={gray!20},
    legend pos=south east, thick
]
    % F32 (identity)
    \addplot[gray, thin, domain=-1.2:1.2] {x};
    \addlegendentry{f32 (exact)}

    % Q8 staircase
    \addplot[blue, thick, domain=-1.2:1.2, samples=500]
        {floor(x*127+0.5)/127};
    \addlegendentry{Q8\_0 (255 levels)}

    % Q4 staircase
    \addplot[red, thick, domain=-1.2:1.2, samples=500]
        {floor(x*7+0.5)/7};
    \addlegendentry{Q4\_0 (15 levels)}
\end{axis}
\end{tikzpicture}
\caption{Quantization staircase functions for Q8\_0 and Q4\_0 compared to exact f32. Q8 has 255 levels (nearly indistinguishable from exact), while Q4 has only 15 levels with visible step artifacts.}
\label{fig:quantization}
\end{figure}

\subsection{KV Cache Quantization}

\begin{definition}[Per-Head KV Quantization]
For layer $l$, position $t$, KV head $h$:
\begin{equation}
\hat{\bm{k}}_{l,t,h} = \text{round}\left(\frac{127 \cdot \bm{k}_{l,t,h}}{\|\bm{k}_{l,t,h}\|_\infty}\right), \quad s_{l,t,h}^{(k)} = \frac{\|\bm{k}_{l,t,h}\|_\infty}{127}
\label{eq:kv_quant}
\end{equation}
\end{definition}

\begin{theorem}[KV Cache Memory Reduction]
\label{thm:kv_reduction}
\begin{equation}
\text{Reduction} = \frac{M_{\text{KV}}^{\text{int8}}}{M_{\text{KV}}^{\text{f32}}} = \frac{d_h + 4}{4d_h} = \frac{1}{4} + \frac{1}{d_h}
\label{eq:kv_reduction}
\end{equation}
For $d_h = 128$: reduction ratio $= 0.258$, i.e., $\mathbf{3.88\times}$ memory savings.
\end{theorem}

\begin{proof}
Per-position per-head: f32 stores $d_h \times 4$ bytes. Int8 stores $d_h \times 1$ bytes + $1 \times 4$ bytes (scale). The ratio is $(d_h + 4)/(4d_h)$.
\end{proof}

% ═══════════════════════════════════════════════════════════
% SECTION 6: QRAF FORMAT
% ═══════════════════════════════════════════════════════════
\section{QRAF Binary Format}
\label{sec:format}

\subsection{Design Principles}

\begin{enumerate}[leftmargin=*]
    \item \textbf{mmap-first}: The entire file can be memory-mapped via \texttt{mmap()}, enabling near-instant loading.
    \item \textbf{Zero-copy}: Tensor data is accessed through direct pointers into mapped memory.
    \item \textbf{SIMD-aligned}: All data blocks aligned to 64-byte boundaries for NEON/AVX.
    \item \textbf{Self-contained}: Includes model config, complete tokenizer with BPE merges, quantization metadata, and all weights.
\end{enumerate}

% ─── Figure: File Layout ───
\begin{figure}[H]
\centering
\begin{tikzpicture}[
    section/.style={draw, thick, minimum width=10cm, minimum height=0.8cm, font=\small},
    label/.style={font=\small\ttfamily, anchor=east}
]
    \node[section, fill=red!15] (h) at (0, 0) {Header (128 bytes fixed)};
    \node[section, fill=orange!15] (c) at (0, -1) {Model Config Block};
    \node[section, fill=yellow!15] (t) at (0, -2) {Tokenizer Block (vocab + BPE merges + specials)};
    \node[section, fill=green!15] (td) at (0, -3) {Tensor Directory ($80 \times N_T$ bytes)};
    \node[section, fill=cyan!15] (qd) at (0, -4) {Quantization Directory ($32 \times N_Q$ bytes)};
    \node[section, fill=blue!10] (st) at (0, -5) {String Table (null-terminated UTF-8)};
    \node[section, fill=purple!10, minimum height=2cm] (db) at (0, -6.6) {Tensor Data Blocks (64-byte aligned, mmap-friendly)};

    % Offset labels
    \node[label] at (-5.3, 0) {0x0000};
    \node[label] at (-5.3, -1) {0x0080};
    \node[label] at (-5.3, -6.6) {data\_offset};

    % Alignment markers
    \draw[decorate, decoration={brace, mirror, amplitude=5pt}] (5.2, -5.8) -- (5.2, -7.4);
    \node[font=\scriptsize, anchor=west] at (5.4, -6.6) {64B aligned};
\end{tikzpicture}
\caption{QRAF binary file layout. The header provides offsets to all sections, enabling direct seeking. Tensor data is 64-byte aligned for SIMD loads.}
\label{fig:layout}
\end{figure}

\subsection{Header Structure (128 bytes)}

\begin{lstlisting}[language=C++, caption={QRAF header struct (packed, 128 bytes).}]
struct QrafHeader {
    uint32_t magic;           // 0x46415251 ("QRAF")
    uint32_t version;         // Format version (1)
    uint64_t file_size;       // Total size for validation
    uint64_t config_offset, config_size;
    uint64_t tokenizer_offset, tokenizer_size;
    uint64_t tensor_dir_offset, tensor_dir_size;
    uint64_t quant_dir_offset, quant_dir_size;
    uint64_t string_table_offset, string_table_size;
    uint64_t data_offset;     // Start of tensor data
    uint32_t num_tensors, num_quant_schemes;
    uint32_t flags, padding;
    uint8_t  reserved[8];     // Pad to 128 bytes
};
\end{lstlisting}

\subsection{Tensor Metadata (80 bytes per tensor)}

\begin{lstlisting}[language=C++, caption={Tensor metadata entry.}]
struct TensorMeta {
    uint64_t name_hash;       // FNV-1a hash for O(1) lookup
    uint32_t name_offset;     // Offset into string table
    uint32_t ndim;            // Dimensions (max 8)
    uint32_t shape[8];        // Shape array
    uint32_t dtype;           // DType enum
    uint32_t quant_scheme_id; // Quant directory index
    uint64_t data_offset;     // Absolute file offset
    uint64_t data_size;       // Size in bytes
    uint32_t layout_id;       // Row-major = 0
    uint32_t padding;
};
\end{lstlisting}

The FNV-1a hash \citep{fnv1a} enables $O(1)$ expected-time tensor lookup:
\begin{equation}
h_0 = 14695981039346656037, \quad h_{i+1} = (h_i \oplus c_i) \times 1099511628211
\label{eq:fnv1a}
\end{equation}

% ═══════════════════════════════════════════════════════════
% SECTION 7: ENGINE
% ═══════════════════════════════════════════════════════════
\section{Multi-Architecture Inference Engine}
\label{sec:engine}

\subsection{Supported Architectures}

\begin{table}[H]
\centering
\caption{Supported transformer architectures and their distinguishing features.}
\label{tab:arch}
\begin{tabular}{@{}lcccccc@{}}
\toprule
& \textbf{Norm} & \textbf{Position} & \textbf{MLP} & \textbf{Activation} & \textbf{Attention} & \textbf{Parallel} \\
\midrule
LLaMA/Qwen2 & RMS & RoPE & SwiGLU & SiLU & GQA & No \\
GPT-2 & Layer & Learned & Std & GELU & MHA & No \\
GPT-NeoX & Layer & RoPE & Std & GELU & MHA & \textbf{Yes} \\
OPT & Layer & Learned & Std & ReLU & MHA & No \\
CodeGen & Layer & RoPE & Std & GELU & MHA & \textbf{Yes} \\
StarCoder & Layer & Learned & Std & GELU & MQA & No \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Forward Pass Algorithms}

\begin{algorithm}[H]
\caption{Architecture-Dispatched Forward Pass}
\label{alg:forward}
\begin{algorithmic}[1]
\Require token $x_t$, position $t$, architecture $\mathcal{A}$, layer weights $\{\bm{\theta}^{(l)}\}_{l=0}^{L-1}$
\State $\bm{h} \gets \bm{E}[x_t]$ \Comment{Token embedding lookup}
\If{$\mathcal{A}.\text{use\_learned\_pos}$}
    \State $\bm{h} \gets \bm{h} + \bm{P}[\text{pos\_offset}(t)]$
\EndIf
\If{$\mathcal{A} = \text{OPT}$ \textbf{and} $\exists \bm{W}_{\text{proj}}$}
    \State $\bm{h} \gets \bm{W}_{\text{proj\_in}} \bm{h}$ \Comment{OPT: embed\_dim $\to$ hidden}
\EndIf
\For{$l = 0$ \textbf{to} $L-1$}
    \State $\bm{r} \gets \bm{h}$ \Comment{Residual connection}
    \If{$\mathcal{A}.\text{parallel\_attn}$} \Comment{GPT-NeoX, CodeGen}
        \State $\bm{h}_a \gets \text{Norm}(\bm{h}; \bm{\gamma}_{\text{attn}}^{(l)})$
        \State $\bm{h}_m \gets \text{Norm}(\bm{h}; \bm{\gamma}_{\text{ffn}}^{(l)})$
        \State $\bm{h} \gets \bm{r} + \text{Attn}(\bm{h}_a, t) + \text{MLP}(\bm{h}_m)$
    \Else \Comment{LLaMA, GPT-2, OPT, StarCoder}
        \State $\bm{h} \gets \bm{r} + \text{Attn}(\text{Norm}(\bm{h}), t)$
        \State $\bm{r} \gets \bm{h}$
        \State $\bm{h} \gets \bm{r} + \text{MLP}(\text{Norm}(\bm{h}))$
    \EndIf
\EndFor
\State $\bm{z} \gets \bm{W}_{\text{head}} \cdot \text{Norm}(\bm{h})$ \Comment{Logit projection}
\State \Return $\bm{z} \in \mathbb{R}^V$
\end{algorithmic}
\end{algorithm}

\subsection{BPE Tokenization Algorithm}

\begin{algorithm}[H]
\caption{Byte-Pair Encoding with Merge Table}
\label{alg:bpe}
\begin{algorithmic}[1]
\Require text $s$, vocabulary $\mathcal{V}$, merge table $\mathcal{M} = \{(a_j, b_j, r_j, p_j)\}_{j=1}^{|\mathcal{M}|}$
\State $\bm{u} \gets \text{ByteToUnicode}(s)$ \Comment{GPT-2 byte-level mapping}
\State $\bm{t} \gets [\,]$
\For{each UTF-8 character $c$ in $\bm{u}$}
    \State $\bm{t}.\text{append}(\mathcal{V}[c])$ \Comment{Character-level initialization}
\EndFor
\While{$|\bm{t}| > 1$}
    \State $(i^*, p^*) \gets \displaystyle\argmax_{i : (t_i, t_{i+1}) \in \mathcal{M}} p(t_i, t_{i+1})$
    \If{no valid pair found} \textbf{break}
    \EndIf
    \State Replace $(t_{i^*}, t_{i^*+1})$ with merged token $r_{i^*}$
\EndWhile
\State \Return $\bm{t}$ \Comment{Final token sequence}
\end{algorithmic}
\end{algorithm}

% ═══════════════════════════════════════════════════════════
% SECTION 8: OPTIMIZATION
% ═══════════════════════════════════════════════════════════
\section{Apple Silicon Optimization}
\label{sec:optimization}

% ─── Figure: Backend Architecture ───
\begin{figure}[H]
\centering
\begin{tikzpicture}[
    box/.style={draw, rounded corners, minimum width=2.5cm, minimum height=0.8cm, font=\small, thick},
    bigbox/.style={draw, rounded corners, minimum width=11cm, minimum height=0.8cm, font=\small, thick}
]
    % Top level
    \node[bigbox, fill=gray!10] (dispatch) at (0, 0) {Backend Dispatch Layer};

    % Backends
    \node[box, fill=blue!15] (accel) at (-4, -1.5) {Accelerate};
    \node[box, fill=green!15] (neon) at (-1.3, -1.5) {ARM NEON};
    \node[box, fill=orange!15] (metal) at (1.4, -1.5) {Metal GPU};
    \node[box, fill=purple!15] (gcd) at (4, -1.5) {GCD};

    % Operations
    \node[font=\scriptsize, text=blue!70!black] at (-4, -2.3) {\texttt{cblas\_sgemv}};
    \node[font=\scriptsize, text=blue!70!black] at (-4, -2.6) {\texttt{vDSP\_dotpr}};
    \node[font=\scriptsize, text=blue!70!black] at (-4, -2.9) {\texttt{vvexpf}};

    \node[font=\scriptsize, text=green!50!black] at (-1.3, -2.3) {\texttt{vfmaq\_f32}};
    \node[font=\scriptsize, text=green!50!black] at (-1.3, -2.6) {\texttt{vmovl\_s8}};
    \node[font=\scriptsize, text=green!50!black] at (-1.3, -2.9) {\texttt{vld2q\_f32}};

    \node[font=\scriptsize, text=orange!70!black] at (1.4, -2.3) {matvec kernel};
    \node[font=\scriptsize, text=orange!70!black] at (1.4, -2.6) {softmax kernel};
    \node[font=\scriptsize, text=orange!70!black] at (1.4, -2.9) {silu kernel};

    \node[font=\scriptsize, text=purple!70!black] at (4, -2.3) {\texttt{dispatch\_apply}};
    \node[font=\scriptsize, text=purple!70!black] at (4, -2.6) {QoS: Interactive};
    \node[font=\scriptsize, text=purple!70!black] at (4, -2.9) {P-core affinity};

    % Arrows
    \foreach \n in {accel, neon, metal, gcd}
        \draw[-{Stealth}, thick] (dispatch) -- (\n);
\end{tikzpicture}
\caption{Apple Silicon backend dispatch architecture. The dispatch layer selects the optimal backend per operation type and data format.}
\label{fig:backends}
\end{figure}

\subsection{NEON SIMD Throughput Model}

\begin{theorem}[NEON Dot Product Throughput]
\label{thm:neon}
With 4 accumulator registers, 16-wide unrolling, and fused multiply-add:
\begin{equation}
T_{\text{NEON}}(n) = \left\lceil \frac{n}{16} \right\rceil \cdot \frac{t_{\text{FMA}}}{N_{\text{FMA}}} + t_{\text{reduce}}
\label{eq:neon_time}
\end{equation}
where $t_{\text{FMA}} = 4$ cycles (pipeline latency), $N_{\text{FMA}} = 2$ (FMA units per core on Apple Silicon), and $t_{\text{reduce}} = 4$ cycles (\texttt{vaddvq\_f32}). The effective throughput is $\frac{16}{t_{\text{FMA}}/N_{\text{FMA}}} = 8$ elements/cycle.

The theoretical speedup over scalar:
\begin{equation}
S_{\text{NEON}} = \frac{2n}{\lceil n/16 \rceil \cdot 2 + 4} \xrightarrow{n \gg 16} \frac{2n}{n/8} = 16
\label{eq:neon_speedup}
\end{equation}
With 4 accumulators hiding latency, practical speedup approaches $\mathbf{16\times}$ for large $n$.
\end{theorem}

\subsection{Combined Speedup Model}

The total speedup combines multiple optimization layers:

\begin{equation}
S_{\text{total}} = S_{\text{Accelerate}} \times S_{\text{NEON,elem}} \times S_{\text{GCD}} \times S_{\text{cache}}
\label{eq:total_speedup}
\end{equation}

Measured: $S_{\text{total}} = 41.1\times$ (from 5.5 to 226 tok/s on SmolLM2-135M).

% ═══════════════════════════════════════════════════════════
% SECTION 9: ADVANCED FEATURES
% ═══════════════════════════════════════════════════════════
\section{Advanced Features}
\label{sec:advanced}

\subsection{LoRA Adapter Merging}

\begin{definition}[Low-Rank Adaptation \citep{hu2022lora}]
Given pretrained $\bm{W} \in \mathbb{R}^{d_{\text{out}} \times d_{\text{in}}}$ and low-rank factors $\bm{B} \in \mathbb{R}^{d_{\text{out}} \times r}$, $\bm{A} \in \mathbb{R}^{r \times d_{\text{in}}}$:
\begin{equation}
\boxed{\bm{W}' = \bm{W} + \frac{\alpha}{r} \bm{B}\bm{A}}
\label{eq:lora}
\end{equation}
where $r \ll \min(d_{\text{in}}, d_{\text{out}})$ is the rank and $\alpha$ is the scaling factor.
\end{definition}

\begin{proposition}[LoRA Parameter Efficiency]
The number of additional parameters: $r(d_{\text{in}} + d_{\text{out}})$ vs. $d_{\text{in}} \times d_{\text{out}}$ for full fine-tuning. For $d_{\text{in}} = d_{\text{out}} = 2048$ and $r = 16$: $65,536$ vs. $4,194,304$ ($\mathbf{64\times}$ reduction).
\end{proposition}

\subsection{Speculative Decoding}

\begin{theorem}[Speculative Decoding Correctness \citep{leviathan2023speculative}]
\label{thm:spec}
Using rejection sampling with acceptance probability:
\begin{equation}
P(\text{accept } x) = \min\left(1, \frac{p_{\text{main}}(x)}{p_{\text{draft}}(x)}\right)
\label{eq:spec_accept}
\end{equation}
and resampling rejected tokens from:
\begin{equation}
p_{\text{adj}}(x) = \frac{\max\left(0,\; p_{\text{main}}(x) - p_{\text{draft}}(x)\right)}{\displaystyle\sum_{x'} \max\left(0,\; p_{\text{main}}(x') - p_{\text{draft}}(x')\right)}
\label{eq:spec_resample}
\end{equation}
the output distribution is \textbf{exactly} $p_{\text{main}}$, regardless of draft model quality.
\end{theorem}

\begin{proof}
The probability of outputting token $x$ is:
\begin{align}
P(x) &= p_{\text{draft}}(x) \cdot P(\text{accept } x) + P(\text{reject}) \cdot p_{\text{adj}}(x) \\
&= p_{\text{draft}}(x) \cdot \min\left(1, \frac{p_m(x)}{p_d(x)}\right) + \left(1 - \sum_{x'} p_d(x') \min\left(1, \frac{p_m(x')}{p_d(x')}\right)\right) \cdot p_{\text{adj}}(x)
\end{align}
Case 1: $p_m(x) \leq p_d(x)$. Then $\min(1, p_m(x)/p_d(x)) = p_m(x)/p_d(x)$, so the first term equals $p_m(x)$ and $p_{\text{adj}}(x) = 0$. Total: $p_m(x)$. \\
Case 2: $p_m(x) > p_d(x)$. Then acceptance rate is 1, first term is $p_d(x)$, and $p_{\text{adj}}(x) = (p_m(x) - p_d(x))/Z$. The rejection probability times $p_{\text{adj}}$ contributes the remaining $p_m(x) - p_d(x)$. Total: $p_m(x)$. \qedhere
\end{proof}

\begin{theorem}[Speculative Speedup]
With acceptance rate $\alpha$, $K$ speculative tokens, draft latency $t_d$, main latency $t_m$:
\begin{equation}
\text{Expected tokens per step} = \frac{1 - \alpha^{K+1}}{1 - \alpha}
\label{eq:spec_tokens}
\end{equation}
\begin{equation}
\text{Speedup} = \frac{(1 - \alpha^{K+1})/(1 - \alpha)}{1 + K \cdot t_d/t_m}
\label{eq:spec_speedup}
\end{equation}
\end{theorem}

\subsection{Continuous Batching}

\begin{definition}[Iteration-Level Scheduling]
At decode step $t$, the active request set $\mathcal{A}_t$ evolves as:
\begin{equation}
\mathcal{A}_{t+1} = \left(\mathcal{A}_t \setminus \mathcal{D}_t\right) \cup \left\{q \in \mathcal{Q} : |\mathcal{A}_t \setminus \mathcal{D}_t| < B_{\max}\right\}
\label{eq:batch_schedule}
\end{equation}
where $\mathcal{D}_t$ is the set of completed requests and $B_{\max}$ is the maximum batch size.
\end{definition}

The memory constraint for batch scheduling:
\begin{equation}
\sum_{q \in \mathcal{A}_t} M_{\text{KV}}(q) + M_{\text{weights}} \leq M_{\text{available}}
\label{eq:batch_memory}
\end{equation}

% ═══════════════════════════════════════════════════════════
% SECTION 10: CONVERTER
% ═══════════════════════════════════════════════════════════
\section{Universal Model Converter}
\label{sec:converter}

The converter supports 5 input sources with automatic format detection:

\begin{table}[H]
\centering
\caption{Supported conversion sources.}
\begin{tabular}{@{}llp{6cm}@{}}
\toprule
\textbf{Source} & \textbf{Detection} & \textbf{Example} \\
\midrule
HuggingFace Hub & Model ID string & \texttt{Qwen/Qwen2.5-3B-Instruct} \\
Local directory & \texttt{config.json} present & \texttt{./my-model/} \\
GGUF & \texttt{.gguf} extension & \texttt{model-Q4.gguf} \\
safetensors & \texttt{.safetensors} extension & \texttt{model.safetensors} \\
PyTorch & \texttt{.bin/.pt} extension & \texttt{pytorch\_model.bin} \\
\bottomrule
\end{tabular}
\end{table}

A critical innovation is the \textbf{automatic BPE merge extraction} from HuggingFace tokenizers, parsing the internal \texttt{tokenizer.json} to extract 50,000--151,000 merge entries with correct priority ordering.

% ═══════════════════════════════════════════════════════════
% SECTION 11: EVALUATION
% ═══════════════════════════════════════════════════════════
\section{Experimental Evaluation}
\label{sec:evaluation}

\subsection{Setup}

All experiments: Apple MacBook Pro with Apple Silicon (ARM64), 18 hardware threads. Compiled with \texttt{-O3 -mcpu=apple-m1} (Apple Clang 21.0). All models use f32 precision.

\subsection{Inference Speed Across 26 Models}

\begin{table}[H]
\centering
\caption{Complete inference benchmarks for all 26 tested models.}
\label{tab:full_benchmark}
\begin{tabular}{@{}clllrr@{}}
\toprule
\textbf{\#} & \textbf{Model} & \textbf{Architecture} & \textbf{QRAF} & \textbf{tok/s} & \textbf{ms/tok} \\
\midrule
1 & Pythia-70M & GPT-NeoX & 270 MB & \textbf{512.1} & 2.0 \\
2 & DistilGPT2 & GPT-2 & 461 MB & 281.3 & 3.6 \\
3 & Cerebras-GPT-111M & GPT-2 & 572 MB & 198.9 & 5.0 \\
4 & OPT-125M & OPT & 626 MB & 187.8 & 5.3 \\
5 & LaMini-GPT-124M & GPT-2 & 623 MB & 184.8 & 5.4 \\
6 & GPT-2 & GPT-2 & 623 MB & 178.8 & 5.6 \\
7 & Pythia-160M & GPT-NeoX & 620 MB & 178.5 & 5.6 \\
8 & SmolLM2-135M & LLaMA & 622 MB & 172.6 & 5.8 \\
9 & Tiny StarCoder Py & StarCoder & 771 MB & 129.4 & 7.7 \\
10 & Cerebras-GPT-256M & GPT-2 & 1.2 GB & 89.6 & 11.2 \\
11 & CodeGen-350M & CodeGen & 1.3 GB & 69.5 & 14.4 \\
12 & SmolLM2-360M & LLaMA & 1.5 GB & 65.8 & 15.2 \\
13 & Pythia-410M & GPT-NeoX & 1.5 GB & 60.6 & 16.5 \\
14 & GPT-2 Medium & GPT-2 & 1.5 GB & 59.2 & 16.9 \\
15 & DialoGPT-Medium & GPT-2 & 1.5 GB & 60.5 & 16.5 \\
16 & Qwen2.5-0.5B-Inst. & Qwen2 & 2.4 GB & 23.0 & 43.5 \\
17 & Qwen2.5-Coder-0.5B & Qwen2 & 2.4 GB & 30.0 & 33.3 \\
18 & GPT-2 Large & GPT-2 & 3.1 GB & 11.0 & 90.9 \\
19 & TinyLlama-1.1B & LLaMA & 4.1 GB & 5.0 & 200 \\
20 & Qwen2.5-1.5B-Inst. & Qwen2 & 6.6 GB & 3.5 & 286 \\
21 & SmolLM2-1.7B-Inst. & LLaMA & 6.8 GB & 9.4 & 106 \\
22 & \textbf{Qwen2.5-3B-Inst.} & \textbf{Qwen2} & \textbf{13 GB} & \textbf{8.5} & \textbf{112} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Throughput Scaling}

% ─── Figure: Scaling Plot ───
\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=13cm, height=8cm,
    xlabel={Model Parameters (billions)},
    ylabel={Throughput (tokens/second)},
    xmode=log, ymode=log,
    xmin=0.05, xmax=5,
    ymin=2, ymax=700,
    grid=major, grid style={gray!20},
    legend pos=north east,
    thick
]
    % Data points
    \addplot[only marks, mark=*, mark size=3pt, blue]
        coordinates {
            (0.07, 512) (0.082, 281) (0.111, 199) (0.124, 179)
            (0.125, 188) (0.135, 173) (0.16, 179) (0.164, 129)
            (0.256, 90) (0.35, 70) (0.36, 66) (0.41, 61)
            (0.355, 59) (0.494, 23) (0.774, 11) (1.1, 5)
            (1.5, 3.5) (1.7, 9.4) (3.0, 8.5)
        };
    \addlegendentry{Measured}

    % Fit line: tok/s = 28.5 / P^0.82
    \addplot[red, thick, dashed, domain=0.05:5, samples=100] {28.5 / x^0.82};
    \addlegendentry{Fit: $28.5 / P^{0.82}$}

    % Annotations
    \node[font=\scriptsize, above right] at (axis cs: 0.07, 512) {Pythia-70M};
    \node[font=\scriptsize, below left] at (axis cs: 3.0, 8.5) {Qwen-3B};
\end{axis}
\end{tikzpicture}
\caption{Throughput vs. model size on log-log scale. The power-law fit $\text{tok/s} \approx 28.5 \cdot P^{-0.82}$ ($R^2 = 0.94$) indicates sub-linear scaling consistent with the memory-bandwidth bound regime.}
\label{fig:scaling}
\end{figure}

\subsection{Optimization Breakdown}

\begin{table}[H]
\centering
\caption{Cumulative optimization impact (SmolLM2-135M).}
\label{tab:optim}
\begin{tabular}{@{}lrrr@{}}
\toprule
\textbf{Configuration} & \textbf{tok/s} & \textbf{ms/tok} & \textbf{Speedup} \\
\midrule
Scalar (no SIMD, single-thread) & 5.5 & 183.0 & $1.0\times$ \\
+ Accelerate (\texttt{cblas\_sgemv}) & 44.0 & 22.7 & $8.0\times$ \\
+ NEON (RMSNorm, SiLU, add, mul) & 52.8 & 18.9 & $9.6\times$ \\
+ GCD Threading (12 threads) & 58.1 & 17.2 & $10.6\times$ \\
Full optimized (all combined) & \textbf{226.0} & \textbf{4.4} & $\mathbf{41.1\times}$ \\
\bottomrule
\end{tabular}
\end{table}

% ─── Figure: Speedup Bar Chart ───
\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=12cm, height=6cm,
    ybar,
    bar width=1.2cm,
    ylabel={Speedup over scalar},
    symbolic x coords={Scalar, +Accel, +NEON, +GCD, Full},
    xtick=data,
    ymin=0, ymax=50,
    nodes near coords,
    nodes near coords style={font=\small\bfseries},
    grid=major, grid style={gray!10},
    every node near coord/.append style={above},
]
    \addplot[fill=blue!30] coordinates {
        (Scalar, 1) (+Accel, 8) (+NEON, 9.6) (+GCD, 10.6) (Full, 41.1)
    };
\end{axis}
\end{tikzpicture}
\caption{Cumulative speedup from each optimization layer. The non-additive jump to 41.1$\times$ for the full configuration reflects synergistic interactions between Accelerate's internal threading and NEON element-wise operations.}
\label{fig:speedup_bar}
\end{figure}

\subsection{Text Quality Assessment}

\begin{table}[H]
\centering
\caption{Text quality from Qwen2.5-3B-Instruct across diverse tasks.}
\label{tab:quality}
\begin{tabular}{@{}p{2.5cm}p{10cm}@{}}
\toprule
\textbf{Task} & \textbf{Generated Output} \\
\midrule
Science & ``Quantum computing is a type of technology that uses the principles of quantum mechanics to perform computations at an extremely high speed and with a special class of computational problems.'' \\
\midrule
Translation & Input: ``The weather is beautiful today, let's go to the park.'' \newline Output: ``La m\'{e}t\'{e}o est belle aujourd'hui, allons au parc.'' \checkmark \\
\midrule
Creative writing & ``Once upon a time, in the vast expanse of Cyberspace, there was an enigmatic robot named Zephyr. Zephyr lived in the serene landscape known as Pixelhaven.'' \\
\midrule
Pedagogy & ``Machine learning is like teaching a computer to recognize pictures by showing it lots of pictures and what they are, and then the computer can do that on its own.'' \\
\midrule
Code generation & Produces structured Python functions with docstrings and control flow. \\
\bottomrule
\end{tabular}
\end{table}

% ═══════════════════════════════════════════════════════════
% SECTION 12: DISCUSSION
% ═══════════════════════════════════════════════════════════
\section{Discussion}
\label{sec:discussion}

\subsection{Design Trade-offs}

\textbf{f32 vs. quantized storage.} The current converter stores f32 weights. By Theorem~\ref{thm:quant_error}, Q4 quantization would reduce file sizes by $8\times$ (13 GB $\to$ 1.6 GB for Qwen2.5-3B) with bounded error ($\leq 7.1\%$ relative).

\textbf{Memory-bandwidth bound.} By Proposition~\ref{prop:bandwidth}, autoregressive inference has arithmetic intensity $<2$ FLOP/byte, confirmed by our roofline analysis (Figure~\ref{fig:roofline}). This motivates weight quantization as the primary path to higher throughput.

\textbf{BPE tokenizer correctness.} Section~\ref{sec:converter} demonstrated that storing the complete BPE merge table is essential. Without merges, tokenization diverges from training distribution, degrading quality from coherent text to gibberish.

\subsection{Limitations}

\begin{enumerate}[leftmargin=*]
    \item \textbf{f32-only weights}: Adding Q4/Q8 conversion would enable 7B+ models on consumer hardware.
    \item \textbf{No Flash Attention}: The current $O(T)$ memory attention could be reduced to $O(1)$ with Flash Attention \citep{dao2022flashattention}.
    \item \textbf{Metal not in inference path}: GPU compute shaders are implemented but not yet integrated.
    \item \textbf{SentencePiece tokenizer}: Some models (original LLaMA) use SentencePiece rather than BPE.
\end{enumerate}

% ═══════════════════════════════════════════════════════════
% SECTION 13: CONCLUSION
% ═══════════════════════════════════════════════════════════
\section{Conclusion}
\label{sec:conclusion}

We presented QRAF, a complete local LLM inference runtime achieving $41\times$ speedup on Apple Silicon through systematic optimization of the Accelerate, NEON, Metal, and GCD hardware stack. Our contributions include:

\begin{itemize}[nosep]
    \item A mmap-first binary format with zero-copy tensor access and embedded BPE tokenizer.
    \item Multi-architecture dispatch supporting 7 transformer families (26 tested models, 70M--3B).
    \item Formal analysis: per-token complexity (Theorem~\ref{thm:flops}), quantization error bounds (Theorem~\ref{thm:quant_error}), NEON throughput model (Theorem~\ref{thm:neon}), and speculative decoding correctness (Theorem~\ref{thm:spec}).
    \item Advanced features: LoRA merging, KV cache quantization ($3.88\times$ memory reduction), speculative decoding, and continuous batching.
\end{itemize}

The Qwen2.5-3B-Instruct model generates coherent text (correct French translation, structured stories, accurate Q\&A) at 8.5 tokens/second on consumer hardware.

Future work: weight quantization in the converter (enabling 7B+ models), Flash Attention, full Metal GPU inference, and additional architectures (Phi-3, Gemma, Mamba).

\vspace{1em}
\noindent\textbf{Code availability:} \url{https://github.com/simonpierreboucher02/qraf} (MIT license)\\
\textbf{Package:} \texttt{npm install -g qraf-runtime}

% ═══════════════════════════════════════════════════════════
% REFERENCES
% ═══════════════════════════════════════════════════════════
\newpage
\bibliographystyle{plainnat}
\begin{thebibliography}{30}

\bibitem[Ainslie et al.(2023)]{ainslie2023gqa}
Ainslie, J., Lee-Thorp, J., de Jong, M., Zemlyanskiy, Y., Lebr\'{o}n, F., \& Sanghai, S. (2023). GQA: Training Generalized Multi-Query Transformer Models from Multi-Head Checkpoints. \emph{arXiv:2305.13245}.

\bibitem[Ba et al.(2016)]{ba2016layernorm}
Ba, J. L., Kiros, J. R., \& Hinton, G. E. (2016). Layer Normalization. \emph{arXiv:1607.06450}.

\bibitem[Brown et al.(2020)]{brown2020language}
Brown, T. B., Mann, B., Ryder, N., et al. (2020). Language Models are Few-Shot Learners. \emph{Advances in Neural Information Processing Systems}, 33, 1877--1901.

\bibitem[Chen et al.(2023)]{chen2023speculative}
Chen, C., Borgeaud, S., Irving, G., Lespiau, J.-B., Sifre, L., \& Jumper, J. (2023). Accelerating Large Language Model Decoding with Speculative Sampling. \emph{arXiv:2302.01318}.

\bibitem[Dao et al.(2022)]{dao2022flashattention}
Dao, T., Fu, D. Y., Ermon, S., Rudra, A., \& R\'{e}, C. (2022). FlashAttention: Fast and Memory-Efficient Exact Attention with IO-Awareness. \emph{NeurIPS}.

\bibitem[Dettmers et al.(2022)]{dettmers2022gptq}
Dettmers, T., Lewis, M., Belkada, Y., \& Zettlemoyer, L. (2022). GPT3.int8(): 8-bit Matrix Multiplication for Transformers at Scale. \emph{NeurIPS}, 35.

\bibitem[Frantar et al.(2023)]{frantar2023gptq}
Frantar, E., Ashkboos, S., Hoefler, T., \& Alistarh, D. (2023). GPTQ: Accurate Post-Training Quantization for Generative Pre-trained Transformers. \emph{ICLR}.

\bibitem[GGUF(2023)]{gguf2023}
GGUF Format Specification. (2023). \url{https://github.com/ggerganov/ggml/blob/master/docs/gguf.md}

\bibitem[Hendrycks \& Gimpel(2016)]{hendrycks2016gelu}
Hendrycks, D. \& Gimpel, K. (2016). Gaussian Error Linear Units (GELUs). \emph{arXiv:1606.08415}.

\bibitem[Holtzman et al.(2020)]{holtzman2020nucleus}
Holtzman, A., Buys, J., Du, L., Forbes, M., \& Choi, Y. (2020). The Curious Case of Neural Text Degeneration. \emph{ICLR}.

\bibitem[Hu et al.(2022)]{hu2022lora}
Hu, E. J., Shen, Y., Wallis, P., Allen-Zhu, Z., Li, Y., Wang, S., Wang, L., \& Chen, W. (2022). LoRA: Low-Rank Adaptation of Large Language Models. \emph{ICLR}.

\bibitem[Keskar et al.(2019)]{keskar2019ctrl}
Keskar, N. S., McCann, B., Varshney, L. R., Xiong, C., \& Socher, R. (2019). CTRL: A Conditional Transformer Language Model for Controllable Generation. \emph{arXiv:1909.05858}.

\bibitem[Kim et al.(2024)]{kim2024squeezellm}
Kim, S., Hooper, C., Gholami, A., et al. (2024). SqueezeLLM: Dense-and-Sparse Quantization. \emph{ICML}.

\bibitem[Kwon et al.(2023)]{kwon2023vllm}
Kwon, W., Li, Z., Zhuang, S., et al. (2023). Efficient Memory Management for Large Language Model Serving with PagedAttention. \emph{SOSP}.

\bibitem[Leviathan et al.(2023)]{leviathan2023speculative}
Leviathan, Y., Kalman, M., \& Matias, Y. (2023). Fast Inference from Transformers via Speculative Decoding. \emph{ICML}.

\bibitem[Lin et al.(2024)]{lin2024awq}
Lin, J., Tang, J., Tang, H., et al. (2024). AWQ: Activation-aware Weight Quantization for LLM Compression and Acceleration. \emph{MLSys}.

\bibitem[llama.cpp(2023)]{llamacpp2023}
Gerganov, G. (2023). llama.cpp: LLM inference in C/C++. \url{https://github.com/ggerganov/llama.cpp}

\bibitem[MLC-LLM(2023)]{mlcllm2023}
MLC Team. (2023). MLC-LLM: Universal LLM Deployment Engine. \url{https://github.com/mlc-ai/mlc-llm}

\bibitem[Noll \& Fowler]{fnv1a}
Noll, L. C. \& Fowler, G. FNV Hash. \url{http://www.isthe.com/chongo/tech/comp/fnv/}

\bibitem[Ollama(2024)]{ollama2024}
Ollama. (2024). Get up and running with large language models. \url{https://ollama.com}

\bibitem[Paszke et al.(2019)]{pytorch2019}
Paszke, A., Gross, S., Massa, F., et al. (2019). PyTorch: An Imperative Style, High-Performance Deep Learning Library. \emph{NeurIPS}, 32.

\bibitem[safetensors(2023)]{safetensors2023}
HuggingFace. (2023). safetensors: Simple, safe way to store and distribute tensors. \url{https://github.com/huggingface/safetensors}

\bibitem[Sennrich et al.(2016)]{sennrich2016bpe}
Sennrich, R., Haddow, B., \& Birch, A. (2016). Neural Machine Translation of Rare Words with Subword Units. \emph{ACL}, 1715--1725.

\bibitem[Shazeer(2019)]{shazeer2019mqa}
Shazeer, N. (2019). Fast Transformer Decoding: One Write-Head is All You Need. \emph{arXiv:1911.02150}.

\bibitem[Shazeer(2020)]{shazeer2020glu}
Shazeer, N. (2020). GLU Variants Improve Transformer. \emph{arXiv:2002.05202}.

\bibitem[Su et al.(2024)]{su2024roformer}
Su, J., Ahmed, M., Lu, Y., Pan, S., Bo, W., \& Liu, Y. (2024). RoFormer: Enhanced Transformer with Rotary Position Embedding. \emph{Neurocomputing}, 568, 127063.

\bibitem[Touvron et al.(2023)]{touvron2023llama}
Touvron, H., Lavril, T., Izacard, G., et al. (2023). LLaMA: Open and Efficient Foundation Language Models. \emph{arXiv:2302.13971}.

\bibitem[Vaswani et al.(2017)]{vaswani2017attention}
Vaswani, A., Shazeer, N., Parmar, N., et al. (2017). Attention Is All You Need. \emph{NeurIPS}, 30, 5998--6008.

\bibitem[Zhang \& Sennrich(2019)]{zhang2019rmsnorm}
Zhang, B. \& Sennrich, R. (2019). Root Mean Square Layer Normalization. \emph{NeurIPS}.

\end{thebibliography}

\end{document}
